When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-27 doc/paper/vpd_et_paper.tex
at methods

\authors{A. Massmann\affil{1}, P. Gentine\affil{1}, C. Lin\affil{1,2}}

@@

\correspondingauthor{Adam Massmann}{akm2203@columbia.edu}

@@

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and, depending on region, decreased relative humidity \citep{Byrne_2013}. Increases in VPD increase the atmospheric demand for evapotranspirated water \citep{Penman_1948, Monteith_1965}, but also stress plant stomata \citep{Leuning_1990, MEDLYN_2011}. The conflicting effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotraspiration (ET) respondes to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand with limited surface regulation though ecosystem resistance regualtion \citep{Swann_2016}. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

@@

We can use intuition about plant water conservation strategy to hypothesize about ET response to changes in VPD. Plants that evolved to conserve water, such as arid shrubs, should be more likely to reduce ET with increasing VPD, and plants that have evolved or have been engineered to care little about water, such as crops, will be more likely to increase ET with increasing VPD. Atmospheric conditions must matter as well. At the ecosystem scale, there are limits to plant water conservation strategies. As atmospheric demand for water (VPD) increases further, ecosystems should begin to reach their water conservation limits and might not be able to entirely limit ET flux to the atmosphere. At this stage any further increase in VPD will most likely drive a (limited) increase in ET, because the increase in atmospheric demand for water overwhelms the limited plant response to conserve water.

The objective of the present manuscript is to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to drive intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In order to quantify plant response to perturbations in VPD, we apply a Penman-Monteith framework to derive theoretical response of ET to VPD. The model is then validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types. Section 2 describes the data used and derives the framework. Section 4 presents results and comparison to observations, and  Section 5 discusses the conclusions.

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015 database,  all sites with at least four years of data (data available at https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/). Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used in the stomatal conductance model in the Methods section (Section \ref{methods}) are only published for six plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreed broadleaf forests (EBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 62 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig}.

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./map.pdf}
\caption{Plant functional type and location of sites used in analysis. ***\textbf{This is just  a placeholder for now, I am currently fixing it***}}
\label{map_fig}
 \end{figure}

The purpose of this study is to examine ecosystem response to atmospheric drying focusing on the growing season. To accomplish this, we filter and quality control the data using a similar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the average Gross Primary Productivity (GPP) exceeds 10\% of the observed 95th percentile of GPP for a given site. GPP is calculated using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues with rain interception and sensor saturation at high relative humidity (\cite{MEDLYN_2011}).
\end{itemize}
Additionally, we restrict data to the daytime, which is identified when downwelling shortwave radiation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To further reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa. Timesteps with negative observed GPP or ET are also removed, and we aggregate half hourly data to hourly averages to reduce noise. After these quality control procedures, 342461 upscaled hourly observations remain.

\subsection{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table 1). \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining an optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with an empirical approach, which describes the dependence of $g_s$ to VPD. This resulted in the following model for leaf-scale stomatal conductance was:

\begin{linenomath*}
  \begin{equation}
  g_{l-s} = g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_s}
  \end{equation}
\end{linenomath*}
Where $g_1$ is a leaf-scale ``slope'' parameter, A is the net CO$_2$ assimilation rate, and $c_s$ is the atmospheric CO$_2$ concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope parameter ($g_1$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation}
  g_1 = \sqrt{\frac{3 \, \Gamma^* \, \lambda}{1.6}},\footnote{Note the above expression has units of of (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the ideal gas law.}
  \end{equation}
\end{linenomath*}

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis (without dark respiration), and $\lambda$ is the marginal water cost of leaf carbon. So, $g_1$ is a leaf-scale term reflecting the tradeoff of water for carbon uptake and growth. The higher $g_1$, the more open the stomata and the more they release water in exchange for carbon.

The Medlyn model for stomatal conductance has been shown to behave very well across PFTs, compared to other models from the leaf to ecosystem scale \citep[][** still unsure about citing Zhou here]{Lin_2015, Zhou_2014}. It is adapted at the ecosystem scale by multiplying $g_{l-s}$ by leaf area index (LAI) and converting units to m s$^{-1}$:

\begin{linenomath*}
  \label{medlyn}
  \begin{equation}
  g_s = \text{LAI} \frac{R \,T}{P} \left( g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_s}\right)
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
    \label{uwue}
  \end{equation}
\end{linenomath*}
is relatively constant across time and moisture conditions within a plant functional type (CITE ALSO CHAGJIE ** ??), in non-extreme drought conditions. If, following \citet{Lin_2015}, we approximate $g_0$ as $0$ (i.e. we neglect cuticular and epidermal losses - a reasonable assumption except in very dry conditions), we can use uWUE to remove the $A$ dependence from $g_s$ in a way that makes the PM equation analytically tractable:

\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{uWUE \; ET}{c_s \; \sqrt{VPD}}
    \label{new_g_s}
  \end{equation}
\end{linenomath*}

Note that $uWUE$ is fit on the ecosystem scale as in \citet{Zhou_2015} so GPP in Equation \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancellation of LAI with the addition of uWUE in Equation \ref{medlyn}. Plugging Equation \ref{new_g_s} into Equation \ref{orig_pen} and rearranging gives a new expression for PM ET directly as a function of VPD with no $A$ dependence:

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_s \sqrt{VPD} }{ R* \; 1.6\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}
    \label{et}
  \end{equation}
\end{linenomath*}

@@

\[uWUE \appropto \sqrt{\frac{c_s - \Gamma}{1.6 \lambda}},\]

@@

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_s \sqrt{VPD} }{ R* \; 1.6\; \sigma \; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}
    \label{et_sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
\sigma = - \frac{g_a \gamma c_s \sqrt{VPD} L_v P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p VPD\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})},
    \label{sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; VPD} = \frac{2\; g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_s }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right)
    \label{d_et}
  \end{equation}
\end{linenomath*}

@@

We note one final comment on our derivation which is relevant for drought indices. If we approximate $c_s$ at a global mean CO$_2$ concentration, then the RHS of Equation \ref{et} is fully defined using commonly available weather station data and the constants published in \citet{Zhou_2015} and \citet{Lin_2015}. This makes Equation \ref{et} a useful alternative to PET in drought indices and hydrometeorological analysis for vegetated surfaces. Equation \ref{et} better reflects the physics of water exchange at the land surface and would only require fitting of $uWUE$ and $g_1$, and can also account for changes in CO$_2$ concentration, contrary to typical drought indices such as the Palmer Drought Severity Index (PDSI) \citep{Swann_2016}.

\begin{table}
\caption{Definition of symbols and variables, with citation for calculation if applicable.}
\centering
\begin{tabular}{l c c c}
\hline
 Variable & Description & Units & Citation \\
\hline
$e_s$  & saturation vapor pressure & Pa  & - \\ 
$T$  & temperature  & K & - \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
  $g_a$  & aerodynamic conductance & m s$^{-1}$  & \makecell{\cite{Thom_1977} \\ \cite{Paulson_1970} \\ \cite{Beljaars_1991}} \\
  $\rho_a$  & air density & kg m$^{-3}$  & - \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
  $VPD$  & vapor pressure deficit & Pa  & - \\
  $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
  $g_s$  & stomatal conductance & m s$^{-1}$  & \cite{MEDLYN_2011} \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  & \cite{MEDLYN_2011} \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
  $LAI$ & leaf area index & -& - \\
  $\sigma$ & uncertainty parameter & -& - \\
  $c_s$ & surface CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./joint_vpd_sigma.pdf}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$.}
\label{joint_vpd_sigma}
\end{figure}

@@

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in Equation \ref{sign} will determine whether ET increases or decreases with increasing VPD. If the second term is larger then the plant control dominates and ET decreases with increasing VPD. However, if the first term is larger, then atmospheric demand dominates and ET increases with increasing VPD.

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_s$) and the psychometric constant ($\gamma$) are relatively constant over the dataset considered here so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions  (ZHOU ET AL, LIN ET AL ** ??)\explain{Pierre, which Zhou and Lin citations were you referring to that show that uWUE changes with very dry soil conditions?}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from a decrease in ET, i.e. the threshold $VPD_{crit}$ such as the derivative vanishes $\frac{\partial \; ET}{\partial \; VPD} = 0$:
\small
\begin{linenomath*}
  \begin{equation}
VPD_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + \sqrt{\frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE}\left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right)
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}
\normalsize
We note that $VPD_{crit}$ mostly depends on the PFT parameters $uWUE$ and $g_1$ but only vary weakly with climate as most other parameters related to the environment are nearly constant. The value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes tot he atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation \citep{De_2013, Nikinmaa_2013, Ryan_2014}.

>>>>

\authors{Adam Massmann\affil{1}, Pierre Gentine\affil{1}, Changjie Lin\affil{1,2}}

@@

\correspondingauthor{Adam Massmann}{206-919-1364; akm2203@columbia.edu}

@@

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and, depending on region, decreased relative humidity \citep{Byrne_2013}. Increases in VPD increase the atmospheric demand for evapotranspirated water \citep{Penman_1948, Monteith_1965}, but also stress plant stomata \citep{Leuning_1990, MEDLYN_2011}.

The opposing effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotraspiration (ET) respondes to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand and fails to account for ecosystem response \citep{Swann_2016}. In reality, plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

@@

We can use intuition about plant water conservation strategy to hypothesize about ET response to changes in VPD. Plants that evolved to conserve water, such as arid shrubs, should be more likely to reduce ET with increasing VPD, and plants that have evolved or have been engineered to care little about water, such as crops, will be more likely to increase ET with increasing VPD. Atmospheric conditions must matter as well. At the ecosystem scale, there are limits to plant water conservation strategies. As atmospheric demand for water (VPD) increases, ecosystems should begin to reach their water conservation limits and might not be able to entirely limit ET flux to the atmosphere. At this stage any further increase in VPD will most likely drive a (limited) increase in ET, because the increase in atmospheric demand for water overwhelms the limited plant response to conserve water.

The objective of the present manuscript is to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to drive intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In order to quantify plant response to perturbations in VPD, we apply a Penman-Monteith framework to derive the theoretical response of ET to VPD. The model is then validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types.

\section{Materials and Methods}
\label{methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015 database,  all sites with at least four years of data (data available at \sloppy https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy). Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used in the stomatal conductance model in the Methods section (Section \ref{methods}) are only published for six plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreed broadleaf forests (EBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 62 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig}.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./map.pdf}
\caption{Plant functional type and location of sites used in analysis. ***\textbf{This is just  a placeholder for now, I am currently fixing it***}}
\label{map_fig}
 \end{figure}

The purpose of this study is to examine ecosystem response to atmospheric drying focusing on the growing season. To accomplish this, we filter and quality control the data using a similar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the average Gross Primary Productivity (GPP) exceeds 10\% of the observed 95th percentile of GPP for a given site. GPP is calculated using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues with rain interception and sensor saturation at high relative humidity (\cite{MEDLYN_2011}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime, which is identified when downwelling shortwave radiation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa, and to reduce errors at low windspeeds we remove all periods with wind magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed GPP or ET are also removed, and we aggregate half hourly data to hourly averages to reduce noise \citep{Lin_2018}. After these quality control procedures, 335,939 upscaled hourly observations remain.

\subsection{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948, Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table 1).

\begin{table}
\caption{Definition of symbols and variables, with citation for calculation if applicable.}
\centering
\small
\begin{tabular}{l c c c}
\hline
 Variable & Description & Units & Citation \\
\hline
$e_s$  & saturation vapor pressure & Pa  & - \\ 
  $T$  & temperature  & K & - \\
  $P$  & pressure & Pa  & - \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
  $g_a$  & aerodynamic conductance & m s$^{-1}$  & \makecell{\cite{Thom_1977} \\ \cite{Paulson_1970} \\ \cite{Beljaars_1991}} \\
  $\rho_a$  & air density & kg m$^{-3}$  & - \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
  $VPD$  & vapor pressure deficit & Pa  & - \\
  $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
  $g_s$  & stomatal conductance & m s$^{-1}$  & \cite{MEDLYN_2011} \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  & \cite{MEDLYN_2011} \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
  $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
  $LAI$ & leaf area index & -& - \\
  $\sigma$ & uncertainty parameter & -& - \\
  $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
  $\lambda$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
  $\Gamma$ & CO$_2$ compensation point & - & - \\
  $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\

\hline
\end{tabular}
\end{table}

\citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining an optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with an empirical approach, which describes the dependence of $g_s$ to VPD. This resulted in the following model for leaf-scale stomatal conductance:

\begin{linenomath*}
  \begin{equation}
  g_{l-s} = g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_a}
  \end{equation}
\end{linenomath*}
Where $g_1$ is a leaf-scale ``slope'' parameter, A is the net CO$_2$ assimilation rate, and $c_a$ is the atmospheric CO$_2$ concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope parameter ($g_1$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation}
  g_1 = \sqrt{\frac{3 \, \Gamma^* \, \lambda}{1.6}},\footnote{Note this expression has units of of (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the ideal gas law.}
  \end{equation}
\end{linenomath*}

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis (without dark respiration), and $\lambda$ is the marginal water cost of leaf carbon ($\frac{\partial \; \text{transpiration}}{\partial \; A}$). So, $g_1$ is a leaf-scale term reflecting the tradeoff of water for carbon uptake and growth. The higher $g_1$, the more open the stomata and the more they release water in exchange for carbon.

The Medlyn model for stomatal conductance has been shown to behave very well across PFTs \citep[][]{Lin_2015}, and can be  adapted to the ecosystem scale by multiplying $g_{l-s}$ by leaf area index (LAI) and converting units to m s$^{-1}$ with the ideal gas law:

\begin{linenomath*}
  \begin{equation}
    g_s = \text{LAI} \frac{R \,T}{P} \left( g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_a}\right)
      \label{medlyn}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
    \label{uwue}
  \end{equation}
\end{linenomath*}
is relatively constant across time and moisture conditions within a plant functional type, in non-extreme drought conditions. If, following \citet{Lin_2015}, we approximate $g_0$ as $0$ (i.e. we neglect cuticular and epidermal losses - a reasonable assumption except in very dry conditions), we can use uWUE to remove the $A$ dependence from $g_s$ in a way that makes the PM equation analytically tractable:

\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{uWUE \; ET}{c_a \; \sqrt{VPD}}
    \label{new_g_s}
  \end{equation}
\end{linenomath*}

Note that $uWUE$ is fit on the ecosystem scale as in \citet{Zhou_2015}, so GPP in Equation \ref{uwue} is really $A \cdot \text{LAI}$. This leads to the cancellation of LAI in the step from Equation \ref{medlyn} to Equation \ref{new_g_s}. Plugging Equation \ref{new_g_s} into Equation \ref{orig_pen} and rearranging gives a new explicit expression for PM, in which dependence on $A$ is removed:

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_a \sqrt{VPD} }{ R* \; 1.6\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}
    \label{et}
  \end{equation}
\end{linenomath*}

@@

\[uWUE \appropto \sqrt{\frac{c_a - \Gamma}{1.6 \lambda}},\]

@@

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_a \sqrt{VPD} }{ R* \; 1.6\; \sigma \; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}
    \label{et_sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
\sigma = - \frac{g_a \gamma c_a \sqrt{VPD} L_v P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p VPD\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})},
    \label{sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; VPD} = \frac{2\; g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_a }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right)
    \label{d_et}
  \end{equation}
\end{linenomath*}

@@

We note one final comment on our derivation which is relevant for drought indices. If we approximate $c_a$ at a global mean CO$_2$ concentration, then the RHS of Equation \ref{et} is fully defined using commonly available weather station data and the constants published in \citet{Zhou_2015} and \citet{Lin_2015}. This makes Equation \ref{et} a useful alternative to PET in drought indices and hydrometeorological analysis for vegetated surfaces. Equation \ref{et} better reflects the physics of water exchange at the land surface and would only require fitting of $uWUE$ and $g_1$, and can also account for changes in CO$_2$ concentration, contrary to typical drought indices such as the Palmer Drought Severity Index (PDSI) \citep{Swann_2016}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{./joint_vpd_sigma.pdf}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$.}
\label{joint_vpd_sigma}
\end{figure}

@@

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_a }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in Equation \ref{sign} will determine whether ET increases or decreases with increasing VPD. If the second term is larger then the plant control dominates and ET decreases with increasing VPD. However, if the first term is larger, then atmospheric demand dominates and ET increases with increasing VPD.

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_a$) and the psychometric constant ($\gamma$) are relatively constant over the dataset considered here so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions  (ZHOU ET AL, LIN ET AL ** ??)\explain{Pierre, which Zhou and Lin citations were you referring to that show that uWUE changes with very dry soil conditions?}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from a decrease in ET, i.e. the threshold $VPD_{crit}$ such as the derivative vanishes $\frac{\partial \; ET}{\partial \; VPD} = 0$:
\small
\begin{linenomath*}
  \begin{equation}
VPD_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_a}{1.6\; R \; \overline{\sigma} uWUE} + \sqrt{\frac{\gamma c_a}{1.6\; R \; \overline{\sigma} uWUE}\left( \frac{\gamma c_a}{1.6\; R \; \overline{\sigma} uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right)
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}
\normalsize
We note that $VPD_{crit}$ mostly depends on the PFT parameters $uWUE$ and $g_1$ but only vary weakly with climate as most other parameters related to the environment are nearly constant. The value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes tot he atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation \citep{De_2013, Nikinmaa_2013, Ryan_2014}.
